Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-07-15 jso.tex
proofread

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index --~smoothness and
squintability~-- that capture the complexities inherent in PP
optimisation problems. These two metrics are evaluated along with JSO
hyper-parameters to determine their effects on JSO success rate. Our
numerical results confirm the positive impact of these metrics on the
JSO success rate, with squintability being the most significant. The JSO
algorithm has been implemented in the \texttt{tourr} package and
functions to calculate smoothness and squintability of indexes are
available in the \texttt{ferrn} package.
\end{abstract}

@@

Projection Pursuit (PP) (\citet{kr69}, \citet{FT74}) is a dimension
reduction technique aimed at identifying informative linear projections
of data. The method involves optimising a objective function known as
the PP index (e.g. \citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data \(X \in \mathbf{R}^{n\times p}\) and an
index function \(f(\cdot)\), PP finds the orthonormal basis
\(A \in \mathbf{R}^{p \times d}\) that maximises the index value of the
projection, \(Y = XA\):

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieving a globally optimal viewpoint of the data.

Significant efforts have been devoted to enhancing the functionality of
PP, improving optimisation (e.g. \citet{posse1995},
\citet{marie-sainte2010}, \citet{grochowski2011}) and extending its use
to visually explore high-dimensional data. \citet{cook1995grand}
introduced the PP guided tour, which monitors the optimisation visually
to see the data projections leading in and out of the optima. An
implementation is available in the R \citep{R} package \texttt{tourr}
\citep{tourr}. \citet{RJ-2021-105} illustrated how to diagnose
optimisation processes, particularly focusing on the guided tour, and
revealed a need for improved optimisation. While improving the quality
of the optimisation solutions in the tour is essential, it is also
important to be able to view the data projections as the optimisation
progresses. Integrating the guided tour with a global optimisation
algorithm that is efficient in finding the global optimal and enables
viewing of the projected data during the exploration process is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO
\citep{chou_novel_2021} -\citep{rajwar_exhaustive_2023}, inspired by the
search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO a promising candidate for enhancing PP
optimisation.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on \citet{barnett1981interpreting}'s
squint angle) and smoothness as introduced in \citet{laa_using_2020}.
Here we mathematically define metrics for squintability and smoothness,
which is new for the field. A series of simulation experiments using
specially structured datasets and various PP indexes are conducted to
assess the behaviour of JSO under different settings and its sensitivity
to hyper-parameter choices (number of jellyfishes and maximum number of
tries). The relationship between the JSO performance, hyper-parameter
tuning and properties of PP indexes is analysed to provide guidance for
practitioners using projection pursuit, and developers of new indexes.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes. Section~\ref{sec-sim-deets} details
two simulation experiments to assess JSO's performance: one comparing
JSO's performance improvements relative to an existing optimiser,
Creeping Random Search (CRS), and the other studying the impact of
different PP index properties on optimisation performance.
Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{itemize}
\tightlist
\item
  \texttt{search\_geodesic()}: provides a pseudo-derivative
  optimisation. It searches locally for the best direction, based on
  differencing the index values for very close projections. Then it
  follows the direction along the geodesic path between planes, stopping
  when the next index value fails to increase.
\item
  \texttt{search\_better()}: is a brute-force optimisation searching
  randomly for projections with higher index values.
\item
  \texttt{search\_better\_random()}: is essentially simulated annealing
  \citep{Bertsimas93} where the search space is reduced as the
  optimisation progresses.
\item
  \texttt{search\_posse()}: implements the algorithm described in
  \citet{posse95}.
\item
  \texttt{search\_polish()}: is a very localised search, to take tiny
  steps to get closer to the local maximum.
\end{itemize}

There are several PP index functions available: \texttt{holes()} and
\texttt{cmass()} \citep{cook1993projection}; \texttt{lda\_pp()}
\citep{lee2005projection}; \texttt{pda\_pp()} \citep{lee2010projection};
\texttt{dcor2d()} and \texttt{splines2d()} \citep{Grimm2016};
\texttt{norm\_bin()} and \texttt{norm\_kol()} \citep{huber85};
\texttt{slice\_index()} \citep{Laa:2020wkm}. Most are relatively simply
defined, for any projection dimension, and implemented because they are
relatively easy to optimise. A goal is to be able to incorporate more
complex PP indexes, for example based on scagnostics (\citet{scag},
\citet{WW08}).

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space in a way that balances
exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the behaviour of jellyfish to navigate towards the best
solution over iterations \citep{chou_novel_2021}.

To solve the optimisation problem embedded in the PP guided tour, a
starting projection, an index function and the maximum number of
iterations are provided as input. Then, the current projection is
evaluated by the index function. The projection is then moved in a
direction determined by a random factor, influenced by how far along we
are in the optimization process. Occasionally, completely new directions
may be taken like a jellyfish might with ocean currents. A new
projection is accepted if it is an improvement compared to the current
one, rejected otherwise. This process continues and iteratively improves
the projection, until the pre-specified maximum number of trials is
reached.

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, arc=.35mm, breakable, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, colback=white, opacityback=0, left=2mm, toprule=.15mm, opacitybacktitle=0.6, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm]

@@

\textbf{Output}: \texttt{optimized\_projection}

\textbf{Initialize} \texttt{current\_best} as the projection with the
best index value from \texttt{current\_projections}, and
\texttt{current\_idx} as the array of index function values for each
projection in \texttt{current\_projections}

\textbf{for} each trial\_id in 1 to max\_tries \textbf{do}

\begin{quote}
Calculate \(c_t\) based on the current\_idx and max\_trial
\end{quote}

@@

\begin{quote}
Define trend based on the current\_best and current\_projections
\end{quote}

@@

\begin{quote}
For each projection, compare with a random jelly and adjust towards or
away from it (active)
\end{quote}
\end{quote}

@@

\begin{quote}
Determine the projection with the best index value as the new
current\_best
\end{quote}
\end{quote}

\begin{quote}
\textbf{if} trial\_id \(\geq\) max\_trial, print the last best
projection \textbf{exit}
\end{quote}

\textbf{return} the set of projections with the updated current\_best as
the optimized\_projection

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of trials.

\citet{laa_using_2020} has proposed five criteria for assessing
projection pursuit indexes (smoothness, squintability, flexibility,
rotation invariance, and speed). Since not all the properties affects
the execution of the optimisation, the focus here is on the first two
properties, namely \emph{smoothness} (Section~\ref{sec-smoothness}) and
\emph{squintability} (Section~\ref{sec-squintability}), and propose two
metrics to evaluate them.

@@

If a PP index function is evaluated at some random points, as is done at
the initialization stage of the jellyfish search optimization, then
these random index values can be interpreted as a random field, indexed
by a space parameter, namely the random projection angle. This analogy
suggests to use this random training sample to fit a spatial model, a
simple one being a (spatial) Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
function and covariance function. The smoothness property comes into
play in the definition of the covariance function: if a PP index is very
smooth, then two close projection angles should produce close index
values (strong correlation); by contrast, if a PP index is not very
smooth, then two close projection angles might give very different index
values (fast decay of correlations with respect to distance between
angles).

@@

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being \(1/2\),
\(3/2\) and \(5/2\). The parameter \(\nu\), called \emph{smoothness
parameter}, controls the decay of the covariance function. As such, it
is an appropriate measure of smoothness of a random field.

In this context, the parameter \(\nu\) is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as GpGp
\citep{guinness2021gpgp} or ExaGeoStatR \citep{abdulah2023large}, to fit
the hyperparameters of a GP covariance function on data. In this
project, we make use of the GpGp package.

@@

In the existing literature, it is commonly understood that a large
squint angle implies that the objective function value is close to
optimal even when the perfect view to see the structure is far away. A
small squint angle means that the PP index value improves substantially
only when the perfect view is close by. As such, low squintability
implies rapid improvement in the index value when near the perfect view.
For PP, a small squint angle is considered to be undesirable because it
means that the optimiser needs to be very close to be able to ``see''
the optima. Thus, it could be difficult for the optimiser to find the
optima. The mathematical formulation of this intuition is proposed
below:

Let \(A \in \mathbf{R}^{p \times 2}\) be the 2D projection basis and
\(A^*\) the optimal basis achieving the maximum index value for a given
projection pursuit problem defined by the shape to find, data dimension,
and the index function \(f(.)\). The projection distance \(d(A, A^*)\),
is defined as the Frobenius norm \(\lVert . \rVert _F\), between the
squared difference of these two bases:

@@

The index value \(f(XA)\) is defined as described in the introduction.
We now focus our attention to how the index value \(f(XA)\) changes with
respect to the projection distance \(d(A, A^*)\), over the course of the
JSO.

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, should play an important role in the
mathematical definition of squintability. Noticing that \(f\) is
expected to be a decreasing function, the squintability metric is
proposed as follows:

@@

On the one hand, \(-\min_{x}f'(x)\) represents the largest value of the
gradient of \(-f\). On the other hand, \(\underset{x}{\arg\min}f'(x)\)
represents the projection distance at which this largest gradient is
attained. Based on the above discussion, it is expected that these two
values should be both high in the case of high squintability (fast
increase in \(f\) early on), and both low in the case of low
squintability (any substantial increase in \(f\) happens very late,
close to the optimal angle). This suggest that their product
(\ref{eq-squintability}) should provide a sensible measure of
squintability. The multiplicative constant \(4\), which can be deemed
arbitrary, does not change the interpretation of the squintability
metric \(\varsigma\); it is here to adjust the range of values of
\(\varsigma\) and simplify the explicit formula for \(\varsigma\)
obtained later on.

@@

The JSO performance is compared with the existing Creeping Random Search
(CRS) \citep{RJ-2021-105, laa_using_2020} currently used in guided tour,
and explores how JSO behaves with different hyper-parameter (number of
jellyfishes and maximum tries) and data dimension combinations. A second
simulation studies the effect of index properties (smoothness and
squintability), along with JSO hyper-parameters, and data dimension, on
the success rate of the JSO performance.

@@

The performance of JSO is investigated in comparison with the existing
optimiser, CRS, with respect to two hyper-parameters (number of
jellyfishes and maximum number of tries). The performance is measured by
success rate, defined as the proportion of simulations that achieves a
final index value within 0.05 of the best index value found among all 50
simulations. This comparison is based on projection pursuit to find the
pipe shape investigated by \citet{laa_using_2020} using two simple and
well-behaved indexes, the \texttt{holes} index, and the
\texttt{splines2d} index. Fifty simulations are conducted with both JSO
and CRS, in four different data dimensions (\(d = 6, 8, 10, 12\)). JSO
uses 100 jellyfishes with 100 maximum number of tries, while the CRS
allows a maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers accounts for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in \citet{RJ-2021-105} is
used here for assessing JSO, where the design parameters are stored
along with index value, projection basis, random seed, and computation
time.

In this simulation, 50 jellyfishes (\texttt{n\_jellies}) and a maximum
of 50 tries (\texttt{max\_tries}) are used. The full data contains 50
simulations, indexed by the variable \texttt{sim}, with the associated
\texttt{seed}s. Recorded in columns \texttt{basis}, \texttt{index\_val},
\texttt{tries}, \texttt{loop} and \texttt{time}, are the projection
basis, index value, iteration ID, jellyfish ID, and computational time,
respectively.

In the second experiment, 50 simulations are conducted at each
hyper-parameter combination to analyse its effect on the JSO success
rate: number of jellyfishes (20, 50, and 100), and maximum number of
tries (50 and 100). Figure~\ref{fig-success-rate} illustrates the
calculation based on the projection found by the above simulation. Here,
the optimal index value is 0.999. Out of the 50 simulations, 44 achieved
an index value within {[}0.949, 0.999{]}, resulting in a success rate of
0.88.

@@

\caption{\label{fig-success-rate}Illustration of success rate
calculation: The best projections identified by the JSO in 50
simulations are sorted by their index values. The highest index value
found across all simulation is 0.999. Out of the 50 simulations, 44
achieved an index value within 0.05 of the best, resulting in a success
rate of 0.88 (44/50).}

@@

This simulation uses two different data shapes, pipe and a sine wave,
investigated in 6D and 8D spaces with six different PP indexes:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. The two JSO hyper-parameters are
varied. A total of 52 combinations result, comprising of 24 computed on
the pipe data and 28 on the sine-wave data. For each case, JSO is run 50
times to calculate the success rate.

Smoothness and squintability are computed for each run, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability} and as illustrated in
Figure~\ref{fig-smoothness} and Figure~\ref{fig-squintability}.

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. The squintability
measure is then calculated by fitting a four-parameter scaled logistic
function of index value against projection distance, estimated by
non-linear least squares.

A generalised linear model is fitted to the resulting data, using a
binomial family and a logit link function. The data is pre-processed by
1) scaling the JSO hyper-parameters by a factor of 10 for
interpretation, 2) creating a new binary variable \texttt{long\_time} to
indicate cases with an average run time over 30 seconds, and 3)
re-coding the success rate for the \texttt{stringy} index as 0, because
none of the 50 simulations correctly identified the sine-wave shape.

@@

\caption{\label{fig-smoothness}Illustration of steps to calculate
smoothness. For a given projection pursuit problem defined by the shape
to find, data dimension and the index function, sample random bases
given the orthonormality contraint. Calculate the projection distance
and the index value for each random basis. Fit a Gaussian process model
to the index values against the projection distances to obtain the
smoothness measure.}

@@

\caption{\label{fig-squintability}Illustration of steps to calculate
squintability. For a given projection pursuit problem defined by the
shape to find, data dimension and the index function, sample random
bases given the orthonormality and projection distance contraint.
Interpolate the sampled bases to the optimal basis and calculate the
projection distance and the index value for each interpolated basis. Bin
the index value by the projection distance, then claculate the average
index value for each bin. Fit the scaled sigmoid function from Equation
\ref{eq-squintability} to the binned index values against the projection
distance using non-linear least square. Use Equation \ref{eq-parametric}
to calculate the squintability measure from the estimated \(\theta_1\)
to \(\theta_4\).}

@@

\includegraphics{optim_files/figure-pdf/fig-proj-1.pdf}

>>>>

\begin{abstract}
The projection pursuit (PP) guided tour interactively optimises a
criteria function known as the PP index, to explore high-dimensional
data by revealing interesting projections. The optimisation in PP can be
non-trivial, involving non-smooth functions and optima with a small
``squint angle'', detectable only from close proximity. To address these
challenges, this study investigates the performance of a recently
introduced swarm-based algorithm, Jellyfish Search Optimiser (JSO), for
optimising PP indexes. The performance of JSO for visualising data is
evaluated across various hyper-parameter settings and compared with
existing optimisers. Additionally, this work proposes novel methods to
quantify two properties of the PP index --~smoothness and
squintability~-- that capture the complexities inherent in PP
optimisation problems. These two metrics are evaluated along with JSO
hyper-parameters to determine their effects on JSO success rate. Our
numerical results confirm the positive impact of these metrics on the
JSO success rate, with squintability being the most significant. The JSO
algorithm has been implemented in the \texttt{tourr} package and
functions to calculate smoothness and squintability are available in the
\texttt{ferrn} package.
\end{abstract}

@@

Projection Pursuit (PP) (\citet{kr69}, \citet{FT74}) is a dimension
reduction technique aimed at identifying informative linear projections
of data. The method involves optimising an objective function known as
the PP index (e.g. \citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}), which defines the
criteria for what constitutes interesting or informative projections.
Given high-dimensional data \(X \in \mathbf{R}^{n\times p}\) and an
index function \(f(\cdot)\), PP finds the orthonormal basis
\(A \in \mathbf{R}^{p \times d}\) that maximises the index value of the
projection, \(Y = XA\):

@@

The index functions can be non-linear and non-convex, hence an effective
and efficient optimisation procedure is essential to explore the data
landscape and achieve a globally optimal viewpoint of the data.

Optimisation of PP is typically investigated in the literature when new
indexes are proposed \citep{posse1995, marie-sainte2010, grochowski2011}
or when visualisation method are used to track the optimisation process.
\citet{cook1995grand} introduced the PP guided tour, which monitors the
optimisation visually to see the data projections leading in and out of
the optima. An implementation is available in the \texttt{tourr} package
\citep{tourr} in R \citep{R}. \citet{RJ-2021-105} illustrated how to
diagnose optimisation processes, particularly focusing on the guided
tour, and revealed a need for improved optimisation. While improving the
quality of the optimisation solutions in the tour is essential, it is
also important to be able to view the data projections as the
optimisation progresses. Integrating the guided tour with a global
optimisation algorithm that is efficient in finding the global optimal
and enables viewing of the projected data during the exploration process
is a goal.

Here, the potential for a Jellyfish Search Optimiser (JSO) to be
integrated with the projection pursuit guided tour is explored. JSO,
\citep{chou_novel_2021} - \citep{rajwar_exhaustive_2023}, inspired by
the search behaviour of jellyfish in the ocean, is a swarm-based
metaheuristic designed to solve global optimisation problems. Compared
to traditional methods, JSO has demonstrated stronger search ability and
faster convergence, and requires fewer tuning parameters. These
practical advantages make JSO a promising candidate for enhancing PP
optimisation.

The primary goal of the study reported here is to investigate the
performance of JSO in PP optimisation for the guided tour. It is of
interest to assess how quickly and closely the optimiser reaches a gobal
optima, for various PP indexes that may have differing complexities. To
observe the performance of JSO with different types of PP indexes,
metrics are introduced to capture specific properties of the index
including squintability (based on \citet{barnett1981interpreting}'s
squint angle) and smoothness as introduced in \citet{laa_using_2020}.
Here we mathematically define metrics for squintability and smoothness,
which is new for the field. A series of simulation experiments using
various datasets and PP indexes are conducted to assess JSO's behaviour
and its sensitivity to hyper-parameter choices (number of jellyfish and
maximum number of tries). The relationship between the JSO performance,
hyper-parameter choices and properties of PP indexes (smoothness and
squintability) is analysed to provide guidance on selecting optimisers
for practitioners using projection pursuit. Additionally, it aims to
guide the design of new indexes that facilitate easy optimization for PP
researchers.

The paper is structured as follows. Section~\ref{sec-background}
introduces the background of PP guided tour and reviews existing
optimisers and index functions in the literature.
Section~\ref{sec-theory} details JSO and introduces metrics that measure
different properties of PP indexes, smoothness and squintability.
Section~\ref{sec-sim-deets} details two simulation experiments to assess
JSO's performance: one comparing JSO's performance improvements relative
to an existing optimiser, Creeping Random Search (CRS), and the other
studying the impact of different PP index properties on optimisation
performance. Section~\ref{sec-sim-res} presents the results.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

@@

\begin{itemize}
\tightlist
\item
  \texttt{search\_geodesic()}: provides a pseudo-derivative
  optimisation. It searches locally for the best direction, based on
  differencing the index values for very close projections. Then it
  follows the direction along the geodesic path between planes, stopping
  when the next index value fails to increase.
\item
  \texttt{search\_better()}: also known as Creeping Random Search (CRS),
  is a brute-force optimisation searching randomly for projections with
  higher index values.
\item
  \texttt{search\_better\_random()}: is essentially simulated annealing
  \citep{Bertsimas93} where the search space is reduced as the
  optimisation progresses.
\item
  \texttt{search\_posse()}: implements the algorithm described in
  \citet{posse95}.
\item
  \texttt{search\_polish()}: is a very localised search, to take tiny
  steps to get closer to the local maximum.
\end{itemize}

There are several PP index functions available: \texttt{holes()} and
\texttt{cmass()} \citep{cook1993projection}; \texttt{lda\_pp()}
\citep{lee2005projection}; \texttt{pda\_pp()} \citep{lee2010projection};
\texttt{dcor2d()} and \texttt{splines2d()} \citep{Grimm2016};
\texttt{norm\_bin()} and \texttt{norm\_kol()} \citep{huber85};
\texttt{slice\_index()} \citep{Laa:2020wkm}. Most are relatively simply
defined, for any projection dimension, and implemented because they are
relatively easy to optimise. A goal is to be able to incorporate more
complex PP indexes, for example, based on scagnostics (\citet{scag},
\citet{WW08}).

@@

JSO mimics the natural movements of jellyfish, which include passive and
active motions driven by ocean currents and their swimming patterns,
respectively. In the context of optimization, these movements are
abstracted to explore the search space, aiming to balance exploration
(searching new areas) and exploitation (focusing on promising areas).
The algorithm aims to find the optimal solution by adapting the
behaviour of jellyfish to navigate towards the best solution over
iterations \citep{chou_novel_2021}.

To solve the optimisation problem embedded in the PP guided tour, a
starting projection, an index function, the number of jellyfish, and the
maximum number of trials (tries) are provided as input. Then, the
current projection is evaluated by the index function. The projection is
then moved in a direction determined by a random factor, influenced by
how far along we are in the optimisation process. Occasionally,
completely new directions may be taken like a jellyfish might with ocean
currents. A new projection is accepted if it is an improvement compared
to the current one, rejected otherwise. This process continues and
iteratively improves the projection, until the pre-specified maximum
number of trials is reached.

\begin{tcolorbox}[enhanced jigsaw, breakable, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, opacityback=0, arc=.35mm, toptitle=1mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, coltitle=black, rightrule=.15mm, bottomrule=.15mm, colback=white]

@@

\textbf{Output}: \texttt{optimised\_projection}

\textbf{Initialize} \texttt{current\_best} as the projection with the
best index value from \texttt{current\_projections}, and
\texttt{current\_idx} as the array of index values for each projection
in \texttt{current\_projections}

\textbf{for} each \texttt{trial\_id} in 1 to \texttt{max\_tries}
\textbf{do}

\begin{quote}
Calculate the time control value, \(c_t\), based on
\texttt{current\_idx} and \texttt{max\_trial}
\end{quote}

@@

\begin{quote}
Define trend based on the \texttt{current\_best} and
\texttt{current\_projections}
\end{quote}

@@

\begin{quote}
For each projection, compare with a random jellyfish and adjust towards
or away from it (active)
\end{quote}
\end{quote}

@@

\begin{quote}
Determine the projection with the best index value as the new
\texttt{current\_best}
\end{quote}
\end{quote}

\begin{quote}
\textbf{if} \texttt{trial\_id} \(\ge\) \texttt{max\_trial}, print the
last best projection \textbf{exit}
\end{quote}

\textbf{return} the set of projections with the updated
\texttt{current\_best} as the \texttt{optimised\_projection}

@@

The JSO implementation involves several key parameters that control its
search process in optimization problems. These parameters are designed
to guide the exploration and exploitation phases of the algorithm. While
the specific implementation details can vary depending on the version of
the algorithm or its application, the focus is on two main parameters
that are most relevant to our application: the number of jellyfish and
the maximum number of tries.

\citet{laa_using_2020} has proposed five criteria for assessing
projection pursuit indexes (smoothness, squintability, flexibility,
rotation invariance, and speed). Since not all index properties affect
the optimisation process, the focus here is on the first two properties,
\emph{smoothness} (Section~\ref{sec-smoothness}) and
\emph{squintability} (Section~\ref{sec-squintability}), for which
metrics are proposed to quantify them.

@@

If a PP index function is evaluated at some random points, as is done at
the initialization stage of JSO, then these random index values can be
interpreted as a random field, indexed by a space parameter, namely the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

The distribution of a Gaussian process is fully determined by its mean
and covariance function. The smoothness property comes into play in the
definition of the covariance function: if a PP index is very smooth,
then two close projection angles should produce close index values
(strong correlation); by contrast, if a PP index is not very smooth,
then two close projection angles might give very different index values
(fast decay of correlations with respect to distance between angles).

@@

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being
\(\frac{1}{2}\), \(\frac{3}{2}\) and \(\frac{5}{2}\). The parameter
\(\nu\), called \emph{smoothness parameter}, controls the decay of the
covariance function. As such, it is an appropriate measure of smoothness
of a random field.

In this context, the parameter \(\nu\) is suggested as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as done at the initialization stage of the jellyfish
search optimization. There exist several R packages, such as
\texttt{GpGp} \citep{guinness2021gpgp} or \texttt{ExaGeoStatR}
\citep{abdulah2023large}, to fit the hyperparameters of a GP covariance
function on data. In this project, the \texttt{GpGp} package is used.

@@

From \citet{barnett1981interpreting} and \citet{laa_using_2020}, a large
squint angle implies that the objective function value is close to
optimal even when the perfect view to see the structure is far away. A
small squint angle means that the PP index value improves substantially
only when the perfect view is close by. As such, low squintability
implies rapid improvement in the index value when near the perfect view.
For PP, a small squint angle is considered to be undesirable because it
means that the optimiser needs to be very close to be able to ``see''
the optima. Thus, it could be difficult for the optimiser to find the
optima. The mathematical formulation of this intuition is proposed
below:

Let \(A \in \mathbf{R}^{p \times d}\) be the d-D projection basis and
\(A^*\) be the optimal basis achieving the maximum index value for a
given projection pursuit problem defined by the shape to find, data
dimension, and the index function \(f(.)\). The projection distance
\(d(A, A^*)\), is defined as the Frobenius norm \(\lVert . \rVert _F\),
between the squared difference of these two bases:

@@

We now focus our attention to how the index value \(f(XA)\) changes with
respect to the projection distance \(d(A, A^*)\), over the course of the
JSO.

It is expected that for a PP index with high squint angle, the
optimization (\ref{eq-optimization}) should make substantial progress
early on. Conversely, for a PP index with low squint angle, it might
take a long while for the optimization to make substantial progress, as
the candidate projections would need to be very close to the optimal one
for the structure of the index function to be visible enough to be
amenable to efficient optimization. This observation suggests that the
extreme values of \(f'\) (the ones for which \(f''=0\), assuming that
\(f\) is twice differentiable), and the projection distances for which
these values are attained, are crucial in the mathematical definition of
squintability. Noticing that \(f\) is expected to be a decreasing
function, the squintability metric is proposed as follows:

@@

\(-\min_{x}f'(x)\) represents the largest value of the gradient of
\(-f\) and \(\underset{x}{\arg\min}f'(x)\) represents the projection
distance at which this largest gradient is attained. It is expected that
these two values should be both high in the case of high squintability
(fast increase in \(f\) early on), and both low in the case of low
squintability (any substantial increase in \(f\) happens very late,
close to the optimal angle). This suggest that their product
(\ref{eq-squintability}) should provide a sensible measure of
squintability. The multiplicative constant \(4\), which can be deemed
arbitrary, does not change the interpretation of the squintability
metric \(\varsigma\); it is here to adjust the range of values of
\(\varsigma\) and simplify the explicit formula for \(\varsigma\)
obtained later on.

@@

The JSO performance is compared with an existing optimiser, Creeping
Random Search (CRS) \citep{RJ-2021-105, laa_using_2020} used in the PP
guided tour to explore JSO's behaviour under different hyper-parameter
and data dimension combinations. The second simulation studies the
effect of index properties (smoothness and squintability), along with
JSO hyper-parameters, and data dimension, on the success rate of the JSO
performance. This section describes the simulation details, with the
results deferred to Section Section~\ref{sec-sim-res}.

@@

he performance of JSO is investigated both in comparison to the existing
optimizer, CRS, and across various hyper-parameter values. The
performance is measured by the success rate, defined as the proportion
of simulations that achieves a final index value within 0.05 of the best
index value found among all 50 simulations (see
Figure~\ref{fig-success-rate} for an illustration). This comparison is
based on projection pursuit to find the pipe shape investigated by
\citet{laa_using_2020} using the \texttt{holes} index.

Fifty simulations are conducted with both JSO and CRS, in four different
data dimensions (\(d = 6, 8, 10, 12\)). JSO uses 100 jellyfishes with a
maximum of 100 tries, while the CRS allows a maximum of 1000 samples at
each iteration before the algorithm terminates. The different numbers
account for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in \citet{RJ-2021-105} for assessing JSO, where the
design parameters are stored along with index value, projection basis,
random seed, and computation time.

Fifty additional simulations are conducted for each hyper-parameter
combination to analyze how they affect the JSO success rate. This
includes variations in the number of jellyfish (20, 50, and 100) and the
maximum number of tries (50 and 100).

@@

\caption{\label{fig-success-rate}Illustration of success rate
calculation: Final projections based on projection pursuit to find the
pipe shape in 8D data using the holes index, optimised by CRS, in 50
simulations. The 50 final projections are sorted by their index values.
The highest index value found across all simulations is 0.969. Out of
the 50 simulations, 43 achieved an index value within 0.05 of the best,
resulting in a success rate of 0.86 (43/50).}

@@

To assess JSO's performance across various scenarios, two different data
shapes, pipe and a sine wave, are investigated in 6D and 8D spaces using
six different PP indexes: \texttt{dcor2d\_2}, \texttt{loess2d},
\texttt{MIC}, \texttt{TIC}, \texttt{spline}, and \texttt{stringy}, with
varied JSO hyper-parameters. A total of 52 combinations result,
comprising of 24 computed on the pipe data and 28 on the sine-wave data.
Again, JSO is run 50 times to calculate the success rate for each
projection pursuit.

Smoothness and squintability are computed following the procedures
outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability} and as illustrated in
Figure~\ref{fig-smoothness} and Figure~\ref{fig-squintability}.

@@

To compute squintability, 50 random bases are sampled and interpolated
to the optimal basis with a step size of 0.005. Index values and
projection distances are calculated for these interpolated bases and the
index values are averaged with a bin width of 0.005. A four-parameter
scaled logistic function is fitted to the index values against
projection distances, estimated by non-linear least squares. The
squintability measure is then calculated as
\ref{eq-squintability-parametric}.

To construct a relationship among success rate, index properties
(smoothness and squintability), and jellyfish hyper-parameters, a
generalised linear model is fitted using a binomial family and a logit
link function. The data is pre-processed by 1) scaling the JSO
hyper-parameters by a factor of 10 for interpretation, 2) creating a new
binary variable \texttt{long\_time} to indicate cases with an average
run time over 30 seconds, and 3) re-coding the success rate for the
\texttt{stringy} index as 0, because none of the 50 simulations
correctly identified the sine-wave shape.

@@

\caption{\label{fig-smoothness}Illustration of steps to calculate
smoothness. For a given projection pursuit problem defined by the shape
to find, data dimension and the index function, 1) sample random bases
given the orthonormality contraint, 2) calculate the projection distance
and the index value for each random basis, and 3) fit a Gaussian process
model of index values against projection distances to obtain the
smoothness measure.}

@@

\caption{\label{fig-squintability}Illustration of steps to calculate
squintability. For a given projection pursuit problem defined by the
shape to find, data dimension and the index function, 1) sample random
bases given the orthonormality and projection distance contraint, 2)
interpolate the sampled bases to the optimal basis and calculate the
projection distance and the index value for each interpolated basis. 3)
bin the index values by projection distances to obtain the average index
value for each bin, 4) fit the scaled sigmoid function in Equation
\ref{eq-squintability} to the binned index values against projection
distances using non-linear least square, 5) calculate the squintability
measure using Equation \ref{eq-squintability-parametric} with parameters
estimated from the model.}

@@

\includegraphics{jso_files/figure-pdf/fig-proj-1.pdf}
